---
title: "Lab2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(naniar)
library(VIM)
library(skimr)
library(ggfortify)
```

1. Load CA pollution burden data
```{r}
ca_pb <- read_csv(here("data", "ca_pollution_burden.csv"))

ca_dem <- read_csv(here("data","ca_census_demographics_2010.csv"))

```

2. Data cleaning and wrangling
```{r}
#ca_pb
ca_pb_nopct <- ca_pb %>% 
  clean_names() %>% 
  select(-contains("perc")) %>% 
  select(-contains("pctl")) %>% 
  select(-latitude, -longitude)

#ca_dem
ca_dem_clean <- ca_dem %>% 
  clean_names()
```

3. PCA 

Look at pollution burden indicators only
```{r}
ca_pb_subset <- ca_pb_nopct %>% 
  select(ozone:solid_waste, asthma:housing_burden)

```

Use the 'prcomp()' function to do the PCA
```{r, eval = FALSE}

pb_pca <- prcomp(ca_pb_subset, scale = TRUE) 
#Gives an error because there are missings

```

Exploring missingness
```{r}
summary(ca_pb_subset) #Highest is 242 - out of 8000 that's not too many to omit

gg_miss_var(ca_pb_subset) #Shows plot of missings

matrixplot(ca_pb_subset) #Red indicates missing, dark is higher numbers
matrixplot(ca_pb_subset, sortby = "poverty") #Organizes poverty from high to low
```

Remove missings
```{r}
#Remove (complete case deletion) any observation with ANY NA values (AKA removes entire row)

ca_pb_nona <- ca_pb_subset %>% 
  drop_na()

#Summarize
summary(ca_pb_nona)

skim(ca_pb_nona) #Gives a nicer summary

```

Now do PCA:
```{r}
my_ca_PCA <- prcomp(ca_pb_nona, scale = TRUE)
my_ca_PCA 

summary(my_ca_PCA)

biplot(my_ca_PCA) #Kind of messy... 

autoplot(my_ca_PCA, #Autoplot works with ggplot so you can use any of the other customizations
         colour = NA, 
         loadings.label = TRUE,
         loadings.label.size = 3, 
         loadings.label.colour = "black",
         loadings.label.repel = TRUE)

#PC1 and PC2 account for ~34% of the variance (makes sense bc we have so many variables)
#Poverty and education look positively correlated 

```

4. PCA for pollution burden + demographics

Join together ca_dem_clean and ca_pb_nopct
```{r}
#Innerjoin - only keep observations from both dataframes that have a match on the other
#Fulljoin keeps everything whether or not there's a match

ca_df <- ca_dem_clean %>% 
  inner_join(ca_pb_nopct, by = c("census_tract_number" = "census_tract"))

#Now drop NAs by compelte case deletion:
ca_nona <- ca_df %>% 
  drop_na()

```

Make a subset of ca_nona that only includes:
-white_percent
-elderly_65_percent
-pm2_5
-pesticides
-traffic
-asthma
-cardiovascular_disease
-poverty

```{r}
my_sub <- ca_nona %>% 
  select(white_percent, elderly_65_percent, pm2_5, pesticides, traffic, asthma, cardiovascular_disease, poverty)
```

Run PCA
```{r}
my_dem_PCA <- prcomp(my_sub, scale = TRUE)

autoplot(my_dem_PCA, #Autoplot works with ggplot so you can use any of the other customizations
         colour = NA, 
         loadings.label = TRUE,
         loadings.label.size = 3, 
         loadings.label.colour = "black",
         loadings.label.repel = TRUE) +
  scale_y_continuous(limits = c(-0.05,0.05))+
  theme_minimal()

```

